---
title: 'debris_flows_prop'
author: 'Erik Rose'
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{debris_flows_prop}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

### Weighting by Area-Slope Relation

We can fit the area-slope relation to a power law where slope $S$ relates to contributing area $A$ by the function:

$$ S = k_sA^{-\theta}  $$

We can relate logged slope to logged contributing area using a linear model, so that:

$$  log(S) = log(k_sA^{-\theta})  $$
$$  log(S) = log(k_s) - \theta log(A)  $$

where $\theta$ is the slope of a linear model with intercept $log(k_s)$.  Using the estimated $\theta$ for all surveyed creeks, we can back-calculate an individual $k_s$ for each channel node holding $\theta$ constant.

```{r, fig.show='hold'}
library(magrittr)
library(muddier)
library(sp)

sub <- creeks[creeks$GRADIENT < .1 & creeks$contr_area < 3, ]
lslope <- (sub$GRADIENT+.015) %>% log # log slope with offset for negative grads
lcontr <- sub$contr_area %>% log # log contr area
as_mod <- lm(lslope ~ lcontr)  # fit AS to power law
as_theta <- as_mod$coefficients[2]  # pull theta value
(sum_mod <- summary(as_mod))
nrow(creeks)
nrow(sub)

# estimate individual ks values using pulled theta
log_ks <- lslope - as_theta * lcontr
ks <- exp(log_ks)
sub$ks <- ks

# values of ks fall between zero and one and do not need to be normalized
summary(ks)
sub$p_ks <- sub$DebrisFlow / ks

summary(sub$p_ks)

# average cross-sectional depth
sub$fill <- (sub$xsec_area + .01) / (sub$valley_width + .01)

p_ks <- sub$p_ks #/ max(sub$p_ks)

mod_fill <- lm(log(sub$fill) ~ log(p_ks))
fill_theta <- mod_fill$coefficients[2]
summary(mod_fill)

acci <- sub$DebrisFlow - sub$ks
lacci <- log(sub$DebrisFlow) - log(sub$ks)
summary(lm(sub$xsec_area ~ acci))
summary(lm(log(sub$xsec_area+.01) ~ lacci))

summary(lm(sub$valley_width ~ acci))
summary(lm(log(sub$valley_width+.01) ~ lacci))

summary(lm(sub$fill ~ acci))
summary(lm(log(sub$fill+.01) ~ lacci))

nacci <- acci + abs(min(acci))
summary(lm(sub$xsec_area ~ nacci))

plot(acci)

plot(sub$contr_area, sub$xsec_area)
plot(sub$contr_area, sub$valley_width)
plot(sub$contr_area, sub$fill)

ca <- sub$contr_area
xs <- sub$xsec_area
vw <- sub$valley_width
fl <- sub$fill
dp <- sub$DebrisFlow
ks <- sub$ks

summary(lm(log(ca) ~ log(vw)))
summary(lm(vs ~ vw))
summary(lm(vs ~ xs))
summary(lm(vs ~ dp / ks))
summary(lm(log(vs-min(vs)+.01) ~ log(dp) / log(ks)))

vw_mod <- lm(log(xs+.01) ~ log(vw))
vw_theta <- vw_mod$coefficients[2]
vs <- log(xs+.01) - vw_theta * log(vw)
vse <- exp(vs)

ca_mod <- lm(log(ca) ~ log(vw))
ca_theta <- ca_mod$coefficients[2]
cs <- log(ca) - ca_theta * log(vw)
cse <- exp(cs)
# summary(lm(cs ~ log(dp/ks)))
# summary(lm(cse ~ dp + ks))
ai <- dp/ks
# summary(lm(cse ~ ai))
# summary(lm(ai ~ cse))
# summary(lm(vse ~ ai + cse))
# summary(lm(vse ~ dp + ks + cse))
# summary(lm(dp ~ vse + ks + cse))
# summary(lm(vse ~ dp))
# summary(lm(vse ~ dp/ks))
# summary(lm(log(dp) ~ vs))

vse_pred <- predict(lm(vse ~ dp/ks), newdata = data.frame(dp = dp, ks = ks))
sub$vse_pred <- vse_pred


pred <- predict(mod_fill, newdata = data.frame(p_ks = p_ks))
pred <- exp(pred)
sub$pred <- pred
sub$vse <- vse

pred_fill <- log(sub$fill) - fill_theta * log(p_ks)
pred_fill <- exp(pred_fill)
sub$pred_fill <- pred_fill

ord <- sub[order(sub$vse_pred), ]
ord$d_cdf <- to_cdf(sub$vse_pred/sum(sub$vse_pred))
ord$d_pmf <- to_pmf(ord$d_cdf)

ord <- ord[order(sub$vse), ]
ord$h <- to_cdf(sub$vse/sum(sub$vse))

# cdf of d with respect to h
ord$d <- 0
for (i in seq_along(ord$h)) {
  if (i == 1) {
    ord$d[i] <- ord$d_pmf[i]
  }
  if (i > 1) {
    ord$d[i] <- ord$d[i - 1] + ord$d_pmf[i]
  }
}


bin_by_2 <- function(var1, var2, bins = 10)  {
  mat <- matrix(c(var1, var2), ncol = 2)
  mat <- mat[order(var1), ]
  vec_len <- length(var1)
  bin_len <- ceiling(vec_len / bins)
  vec_ids <- 1:length(var1)
  bin_list <- list()
  ranges <- vector(bins, mode = 'numeric')
  xs <- vector(bins, mode = 'numeric')
  ys <- vector(bins, mode = 'numeric')
  for (i in 1:bins)  {
    if (i == 1)  {
      bin_list[[i]] <- vec_ids[vec_ids <= bin_len]
      bin <- mat[bin_list[[i]], ]
      ranges[i] <- bin[nrow(bin), 1]
      xs[i] <- bin[nrow(bin),1] - bin[1,1]
      ys[i] <- bin[nrow(bin),2] - bin[1,2]
    }

    if (i > 1)  {
      bin_list[[i]] <- vec_ids[vec_ids > (i-1) * bin_len &
                                 vec_ids <= i * bin_len]
      bin <- mat[bin_list[[i]], ]
      ranges[i] <- bin[nrow(bin), 1]
      xs[i] <- bin[nrow(bin),1] - bin[1,1]
      ys[i] <- bin[nrow(bin),2] - bin[1,2]
    }
  }
  df <- data.frame(x = xs, y = ys, rng = ranges)
  return(df)
}

df <- bin_by_2(ord$h, ord$d, bins = 27)


wt_h <- df$x / df$y
plot(df$rng, wt_h, xlab = 'F_h', type = 'l')

ord$wt_h <- 0
for (i in seq_along(df$rng)) {
  if (i == 1) {
    ord$wt_h[ord$h <= df$rng[i]] <- wt_h[i]
  }
  if (i > 1) {
    ord$wt_h[ord$h > df$rng[i-1] & ord$h <= df$rng[i]] <- wt_h[i]
  }
}

ord$F_d <- (ord$wt_h * ord$d) / sum(ord$wt_h * ord$d)
ord <- ord[order(ord$F_d), ]
ord$f_d <- to_pmf(ord$F_d)

opt_mod <- lm(ord$vse ~ ord$DebrisFlow / ord$ks * ord$wt_h) 
opt_theta <- opt_mod$coefficients[2]
#log_dO <- log(ord$fill) - opt_theta * log(ord$p_ks) * ord$wt_h
#ord$dO <- exp(log_dO)

#ord$pO <- (ord$dO * ord$ks * opt_theta) / sum(ord$dO * ord$ks * opt_theta)
ord$pO <- (ord$DebrisFlow * ord$wt_h) / sum(ord$DebrisFlow * ord$wt_h)
plot(ord$contr_area, ord$pO)
points(ord$contr_area, ord$DebrisFlow/sum(ord$DebrisFlow), col = 'slateblue', pch = 20)


setwd('/home/crumplecup/work')
png('delivery_probs_adjusted.png',
    width = 7, height = 5, units = 'in', res = 300)
plot(ord$p_ks, ord$pO, pch = 20, log = 'x',
     col = rgb(red=.267, green=.671, blue=.471, alpha=.33), 
     ylab = 'delivery probability', xlab = 'p / k_s',
     ylim = c(min(ord$pO - ord$DebrisFlow/sum(ord$DebrisFlow)), max(ord$pO)))
points(ord$p_ks, ord$DebrisFlow/sum(ord$DebrisFlow),
       pch = 20, col = rgb(red=.306, green=.569, blue=.831, alpha=.33))
points(ord$p_ks, ord$pO - ord$DebrisFlow/sum(ord$DebrisFlow),
       pch = 20, col = rgb(red=1, green=.592, blue=.478, alpha=.33))
legend('bottomright', legend = c('p', 'pO', 'pO-p'), 
       fill = c(rgb(red=.306, green=.569, blue=.831, alpha=.33), 
                rgb(red=.267, green=.671, blue=.471, alpha=.33), 
                rgb(red=1, green=.592, blue=.478, alpha=.33)))
dev.off()

setwd('/home/crumplecup/work')
png('delivery_probs_optimized.png',
    width = 7, height = 5, units = 'in', res = 300)
plot(ord$pO / ord$ks, ord$pO, pch = 20, log = 'x',
     col = rgb(red=.267, green=.671, blue=.471, alpha=.33), 
     ylab = 'delivery probability', xlab = 'pO / k_s',
     ylim = c(min(ord$pO - ord$DebrisFlow/sum(ord$DebrisFlow)), max(ord$pO)))
points(ord$pO / ord$ks, ord$DebrisFlow/sum(ord$DebrisFlow),
       pch = 20, col = rgb(red=.306, green=.569, blue=.831, alpha=.33))
points(ord$pO / ord$ks, ord$pO - ord$DebrisFlow/sum(ord$DebrisFlow),
       pch = 20, col = rgb(red=1, green=.592, blue=.478, alpha=.33))
legend('topleft', legend = c('p', 'pO', 'pO-p'), 
       fill = c(rgb(red=.306, green=.569, blue=.831, alpha=.33), 
                rgb(red=.267, green=.671, blue=.471, alpha=.33), 
                rgb(red=1, green=.592, blue=.478, alpha=.33)))
dev.off()


```


```{r, fig.show='hold'}
library(fitdistrplus)

#subset debris flow deposit ages
df <- as.matrix(char_pmfs)
df <- df[ , charcoal$facies == 'DF']

#determine which sites have multiple samples
site <- charcoal$family[charcoal$facies == 'DF']
counts <- 0
for (i in seq_along(site)) {
  counts[i] <- length(site[site == site[i]])
}
mults <- site[counts > 1] %>% factor %>% levels

#select youngest age from each set of multiples
dfmn <- charcoal$mn[charcoal$facies == 'DF']
mins <- vector(length(site), mode = 'numeric')
for (i in seq_along(mults)) {
  mins[dfmn == min(dfmn[site == mults[i]])] <- 1
}

#screen out all but youngest from multiples
df <- df[ , counts == 1 | (counts > 1 & mins > 0)]

#order from youngest to oldest
dfmn <- dfmn[counts == 1 | (counts > 1 & mins > 0)]
df <- df[ ,order(dfmn)]

# estimate interarrival times
index <- char_pmfs %>% rownames %>% as.numeric # age represented by row
df_n <- length(dfmn)
df_mns <- array(0, c(df_n-1, 1))  # mean interarrival time
df_ar <- array(0, c(df_n-1, nrow(char_pmfs)))  # interarrival age pmf

for(i in 1:(df_n-1)) {
  df_ar[i,] <- convo(df[,i+1], df[,i], index)  # convolved difference in age
  df_mns[i] <- weighted.mean(sort(index), df_ar[i,])
}

# fit exponential distribution to interarrival times
descdist(as.numeric(df_mns))
df_exp <- fitdist(as.numeric(df_mns), "exp")
df_gam <- fitdist(as.numeric(df_mns), 'gamma')
df_wbl <- fitdist(as.numeric(df_mns), 'weibull')

labs <- c('exp', 'gamma', 'weibull')
denscomp(ft = list(df_exp, df_gam, df_wbl), legendtext = labs)
cdfcomp(ft = list(df_exp, df_gam, df_wbl), legendtext = labs)
qqcomp(ft = list(df_exp, df_gam, df_wbl), legendtext = labs)
ppcomp(ft = list(df_exp, df_gam, df_wbl), legendtext = labs)

gofstat(list(df_exp, df_gam, df_wbl))

boot_exp <- bootdist(df_exp, niter =1001, silent = T)
summary(boot_exp)
```












